Biden Expected to Win 54% of Popular Vote in 2020 US Presidential Election

### 2020-11-03 outputs/paper/paper.Rmd
so close

---
title: "Biden Expected to Win 54% of Popular Vote in 2020 US Presidential Election"
author: "James Bao, Zakir Chaudry, Alan Chen, Xinyi Zhang"
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: "Code and data supporting this analysis are available at: https://github.com/JamesBond0014/sta304_ps4."
abstract: |
  | After Trump's shocking upset in the 2016 US Presidential election, political pundits (and American in general) have been closely monitoring the 2020 US presidential election which will determine the leader of the free world for the next four years. In this paper we developed a multi-level regression model with post-stratification by training a model using voter survey results and predicting the outcome of the popular vote using large-scale demographic data for the American population. Our model predicts that Joe Biden will win the 2020 Presidential election with a 53.5% of the popular vote and we established that gender and education are the most significant variables when forecasting voting. Our prediction as well as our breakdown of votes by demographic group could potentially provide the candidates of the election with information on how to target voter.
  |
  | **Keywords:** forecasting, US 2020 Election, Trump, Biden, multilevel regression with post-stratification

@@

Unless you've been living under a rock, the question in everyone minds is who will the 2020 US presidential election. Polls are consistently being updated, attempting to reflect the newest results, as candidates attempts to sway various groups in USA in order to gain more votes. By choosing 5 explanatory variables composed of: age (by groups of 10 years), gender, state, race and education we fit an logistic regression model with multilevel regression with post-stratification in order to predict the outcome of this highly anticipated election.

@@

# Discussion
Figure \@ref(fig:ageproportions) features the trend of the younger voters (between the ages of 18 and 38) to vote left-wing. There are many different reasons this can be the case. Younger people generally tend to be left wing (pew 1). In addition, given the current state of the economy, combined with the fact that in general millennials and Gen-Xers are poorer than their parents, many younger voters are in disillusion with the current government, leading to vote against the Republican Party (mckinsey).

@@

Figure \@ref(fig:genderprop)’s comparison of the votes by gender reveals some interesting findings. While the majority of male voters support Trump, the real highlight of this figure is the fact that nearly 60% of female voters support Biden, essentially twice the difference between Biden and Trump for male voters. Again, there’s many reasons why this could be the case, but there are two major reasons that present themselves.

@@

Overall, while there in majority men vote support Trump, a greater percentage of women support Biden. In conjunction with the metric that women vote more than man, this finding supports a Biden victory.

As discussed in the results, race can also play a significant factor in predicting an individuals vote in the 2020 election. As shown in Figure \@ref(fig:raceproportions), 90% of African American or Black voters are projected to vote for Biden, with Asian voters follow closely with almost 80% of Chinese voters projected to vote for Biden. Given Trump's response to the Black Life's Matter movement, and his response to Coronavirus (China virus according to him), it seems clear that voters belonging to these groups feel strongly against him. There has been protests and outrage country wide in response to police brutality leading to the deaths of black or African American citizens, many protesters feel President Trump is ignoring their voice, focusing on the riots and ignoring the multitude of peaceful protests also happening, going as far as implying the movement is a "symbol of hate (Liptak & Holmes, 2020)." The same article brings up other acts of Trump, such as re-tweeting videos of supporter chanting "White power", Trump diminishing the notion of Black Lives Matter by correcting it into All Lives Matter,  and defending an individual who was accused and convicted of shooting and killing 2 black lives matter protesters (Breuninger, 2020). Googling the words Trump and Black Lives Matter results in pages upon pages of news articles, blog posts and tweets of President Trump seemingly repeatedly condemning Black Lives Matter supporters and supporting anti-black lives matter movements.

Whether or not these articles have taken his words out of context and twisted his words around, it seems clear that Trump is labeled to be "anti-black" and our predictions indicate that these feelings among the Black and African American community will reflect heavily on their votes, with less than 10% projected to be voting for Trump. Another major issue is the global pandemic caused by the SARS-CoV-2 virus (or according to Trump, the China Virus). Asian Americans live in fear of xenophobia and other racist attacks, as many interviewed in the New York Times article (Tavernise & Oppel, 2020) recounts racist attacks and comments, many calling for Trump to stop targetting China. The Vancouver police has reported a 600% increase in reports of hate crimes against the Asian community (Gill, 2020), and while this is not in the USA, it's obvious that Asians living in the USA has decided to hedge their support towards Biden. Of course there are many reasons that affect an individual's decision to vote, but these recent and relevant events may be one of the reasons that voters of Asian and Black or African American races are supporting Joe Biden in the upcoming election. In general, all ethnic minorities (with the exception of American Indian or Alaska Native) support Biden more to various degrees.

In contrast, White and American Indian or Alaska Native groups are predicted to vote more for Trump in comparison to Biden. This is interesting they are different from many other minority groups, who strongly prefer Biden rather than trump. In an article by the CNBC written by Myong (2020), American Indian or Alaska Natives are considered to be a key demographic as they could help swing votes in as many as 7 states. This could explain the why the votes are predicted to be split between Trump and Biden, as both parties are putting a focus on gaining these votes. Similarily with the White population, as they represent the vast majority of citizens in USA, they will be the target race demographic for both Trump and Biden as they both fight for what is predicted to be a close election race.

Figure \@ref(fig:educationprop) outlines the difference in voting based on the voter’s level of education. The first thing of note is that there is a slight majority for Trump for those who have not attended university. However, the difference is so small that it’s essentially equal from a top down perspective. In comparison, those who have attended university tend to support Biden. There are a few reasons this could be the case. One is the notion that universities “make” people more left-wing. Whether or not this is truly the case, and if so for what reason, aside, this is a commonly held belief. However, this may not be the sole reason why such a divide exists. In recent years, conservatives have majorly considered higher education to have a negative effect on the country. (pewsocial) Inferring from this, conservatives are thus less likely to attend university or view university with a particularly positive perspective than their left-wing counterparts.

>>>>

---
title: "Biden Expected to Win 54% of Popular Vote in 2020 US Presidential Election"
subtitle: "4% Confidence Interval Based on a Survey from June 2020"
author: "James Bao, Zakir Chaudry, Alan Chen, Xinyi Zhang"
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: "Code and data supporting this analysis are available at: https://github.com/JamesBond0014/sta304_ps4."
abstract: |
  | After a shocking upset in the 2016 US Presidential election, everyone has their eyes on the 2020 election to determine the leader of the free world for the next four years. In this Paper we trained a model using survey level survey results . Our model predicts that Joe Biden will win the 2020 Presidential election with a 53.5% of the popular vote and a confidence level of 95%. Our prediction as well as our breakdown of votes by demographic group could potentially provide the candidates of the election with information on how to target voters.
  |
  | **Keywords:** forecasting, US 2020 Election, Trump, Biden, multilevel regression with post-stratification

@@

The whole world has eyes on the US 2020 Election, with an intense race between Joe Biden and Donald Trump, the question in everyone minds is who will win. Polls are consistently being updated, attempting to reflect the newest results, as candidates attempts to sway various groups in USA in order to gain more votes. By choosing 5 explanatory variables composed of: age (by groups of 10 years), gender, state, race and education we fit an logistic regression model with multilevel regression with post-stratification in order to predict the outcome of this highly anticipated election.

@@

# Discussion
Figure \@ref(fig:ageprop) features the trend of the younger voters (between the ages of 18 and 38) to vote left-wing. There are many different reasons this can be the case. Younger people generally tend to be left wing (pew 1). In addition, given the current state of the economy, combined with the fact that in general millennials and Gen-Xers are poorer than their parents, many younger voters are in disillusion with the current government, leading to vote against the Republican Party (mckinsey).

@@

Figure 2’s comparison of the votes by gender reveals some interesting findings. While the majority of male voters support Trump, the real highlight of this figure is the fact that nearly 60% of female voters support Biden, essentially twice the difference between Biden and Trump for male voters. Again, there’s many reasons why this could be the case, but there are two major reasons that present themselves.

@@

Overall, while there in majority men vote support Trump, a greater percentage of women support Biden. In conjunction with the metric that women vote more than man, this finding supports a Biden victory. (gender diff)

## Race

As discussed in the results, race can play a significant factor in predicting an individuals vote in the 2020 election. As shown in Figure N, 90% of African American or Black voters are projected to vote for Biden, with Asian voters follow closely with almost 80% of Chinese voters projected to vote for Biden. Given Trump's response to the Black Life's Matter movement, and his response to Coronavirus (China virus according to him), it seems clear that voters belonging to these groups feel strongly against him. There has been protests and outrage country wide in response to police brutality leading to the deaths of black or  African American citizens, many protesters feel President Trump is ignoring their voice, focusing on the riots and ignoring the multitude of peaceful protests also happening, going as far as implying the movement is a "symbol of hate (@trump_blm)." The same article brings up other acts of Trump, such as re-tweeting videos of supporter chanting "White power", Trump diminishing the notion of Black Lives Matter by correcting it into All Lives Matter,  and defending an individual who was accused and convicted of shooting and killing 2 black lives matter protesters (@trump_kyle). Googling the words Trump and Black Lives Matter results in pages upon pages of news articles, blog posts and tweets of President Trump seemingly repeatedly condemning Black Lives Matter supporters and supporting anti-black lives matter movements. Whether or not these articles have taken his words out of context and twisted his words around, it seems clear that Trump is labeled to be "anti-black" and our predictions indicate that these feelings among the Black and African American community will reflect heavily on their votes, with less than 10% projected to be voting for Trump. Another major issue is the global pandemic caused by the SARS-CoV-2 virus (or according to Trump, the China Virus). Asian American's live in fear of xenophobia and other racist attacks, as many interviewed in the New York Times article (@chinese_safety) recounts racist attacks and comments, many calling for Trump to stop targeting China. The Vancouver police has reported a 600% increase in reports of hate crimes against the Asian community, and while this is not in the USA, it's obvious that Asians living in the USA has decided to hedge their support towards Biden. Of course there are many reasons that affect an individual's decision to vote, but these recent and relevant events may be one of the reasons that voters of Asian and Black or African American races are supporting Joe Biden in the upcoming election. In general, all ethnic minorities (with the exception of American Indian or Alaska Native) support Biden more to various degrees.

In contrast, White and American Indian or Alaska Native groups are predicted to vote more for Trump in comparison to Biden. This is interesting they are different from many other minority groups, who strongly prefer Biden rather than trump. In an article by the CNBC written by @native_vote, American Indian or Alaska Natives are considered to be a key demographic as they could help swing votes in as many as 7 states. This could explain the why the votes are predicted to be split between Trump and Biden, as both parties are putting a focus on gaining these votes. Similarily with the White population, as they represent the vast majority of citizens in USA, they will be the target race demographic for both Trump and Biden as they both fight for what is predicted to be a close election race.

## Education

Figure 4 outlines the difference in voting based on the voter’s level of education. The first thing of note is that there is a slight majority for Trump for those who have not attended university. However, the difference is so small that it’s essentially equal from a top down perspective. In comparison, those who have attended university tend to support Biden. There are a few reasons this could be the case. One is the notion that universities “make” people more left-wing. Whether or not this is truly the case, and if so for what reason, aside, this is a commonly held belief. However, this may not be the sole reason why such a divide exists. In recent years, conservatives have majorly considered higher education to have a negative effect on the country. (pewsocial) Inferring from this, conservatives are thus less likely to attend university or view university with a particularly positive perspective than their left-wing counterparts.
